Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-02-01 Manuscript/Figures/figure_legends.Rmd
edits to legend and manuscript

**Figure 9-13. Recursive Feature Elimination for 10 kb data** Beginning with the full set of distance type predictors for each cell line, models were recursively evaluated at different subsets of feature size. For convenience and computational efficiency, feature size decreasing in powers of 2 was considered. For the 10 kb data, near optimal performance was achieved for both the GM12878 (red) and K562 cell lines (blue). The maximum size for feature subsets differed between cell line due the available annotations provided in Supplementary Table 1.

Performances of a series of random forest models with different numbers of the most predictive features for each cell line was compared using AUPRC. For convenience and computational efficiency, only specific subsets of features were considered (increasing powers of 2).

Additionally, models using the full feature space were also evaluated, hence the differing number of features between cell lines. For the 10 kb data, near optimal performance was achieved for both the GM12878 (red) and K562 cell lines (blue).

**Figure 6. Variable importance heatmap for 10 kb data across cell lines.** Predictive importance of the top 32 genomic features across cell lines for 10 kb data. Importance was discretized into quantiles, with features that were not congruent between cell lines filled in grey.

>>>>

**Figure 9-13. Recursive Feature Elimination for 10 kb data** Beginning with the full set of distance type predictors for each cell line, models built on 10 kb resolution data were recursively evaluated at different subsets of feature size. For convenience and computational efficiency, feature size decreasing in powers of 2 was considered. The maximum size for feature subsets differed between cell line due the available annotations that were considered and provided in Supplementary Table 1. Near optimal performance was achieved using the top 32 distance type predictors for both the GM12878 (red) and K562 cell lines (blue).

### 2019-02-01 Manuscript/Supplementary/supplementary_legends.Rmd
edits to legend and manuscript

**Supplementary Figures 9-23. Precision-Recall curves for unbalanced 25 kb, 50kb, and 100kb data.** Precision-recall curves were used to initially compare performance of random forest models with different predictor types on 25 kb, 50kb, and 100kb resolution data, across each resampling technique. GM12878 (solid line) and K562 (dashed line) cell lines were compared with predictor type-specific models which included overlap counts (blue), overlap percents (purple), log distance (green), and all three combined (red). The baseline threshold for each PR curve is defined as the ratio of genomic bins that overlapped with a flanked TAD boundary divided by the total number of genomic bins for the test set, respective to cell line and resolution.

@@

**Figure 3. Model Performances on 25 kb, 50kb, and 100kb data using distance-type predictors for different number of features** Using RFE, performances of a series of random forest models with different numbers of the most predictive features for each cell line was compared using AUPRC. For convenience and computational efficiency, only specific subsets of features were considered (increasing powers of 2). For each resolution, near optimal performance was achieved for both the GM12878 (red) and K562 cell lines (blue).

**Figure 4. Variable importance heatmap for 25 kb, 50kb, and 100kb data across cell lines.** Predictive importance of the top 32 genomic features across cell lines for 25 kb, 50kb, and 100kb resolutions. Importance was discretized into quantiles, with features that were not congruent between cell lines filled in grey.

**Figure 5: Imbalance ratios across resolutions and cell lines.** Imbalance ratios, defined as the ratio of minority class samples to majority class samples, were plotted across resolutions for cell lines GM12878 (red) and K562 (blue). The mean imbalance ratios across resolutions were found to be X1 (SD=Y2) and X2 (SD=Y2) for the GM12878 and K562 cell lines respectively.

>>>>

**Supplementary Figures 9-23. Precision-Recall curves for 25 kb, 50kb, and 100kb data.** Precision-recall curves were used to initially compare performance of random forest models with different predictor types on 25 kb, 50kb, and 100kb resolution data, across each resampling technique. GM12878 (solid line) and K562 (dashed line) cell lines were compared with predictor type-specific models which included overlap counts (blue), overlap percents (purple), log distance (green), and all three combined (red).

@@

**Figure 3. Recursive Feature Elimination for 25 kb, 50 kb, and 100 kb data** Beginning with the full set of distance type predictors for each cell line, models built on 25 kb, 50 kb, and 100 kb resolution data were recursively evaluated at different subsets of feature size. For convenience and computational efficiency, feature size decreasing in powers of 2 was considered. The maximum size for feature subsets differed between cell line due the available annotations that were considered and provided in Supplementary Table 1. Near optimal performance was achieved using the top 32 distance type predictors for both the GM12878 (red) and K562 cell lines (blue).

### 2019-02-01 Manuscript/manuscript.Rmd
edits to legend and manuscript

We evaluated and compared multilpe random forest classification algorithms built on domain data extracted from Hi-C datasets for X cell lines C = {GM12878, K562, ...}, at four different resolutions R = {10 kb, 25 kb, 50 kb, 100 kb}. For each classifier, we compared 3 predictor types P = {OC, OP, Distance} and 5 class balancing techniques B = {None, RUS, ROS, SMOTE, ROSE}. A system produced from the ensemble framework was defined as: E = {c, r, p, b}, where $c \in C, \quad R \in R, \quad p \in P,$ and $b \in$ B. Letting |X| denote the cardinality of each set, we evaluated $|C| \times |R| \times |P| \times |B|$ different ensemble systems, totaling X (=$x \times 4 \times 3 \times 5$) separate models. For simplicity, we present the most significant results, with the complete results available as Supplementary Tables and Figures.

@@

For each cell line, we performed a series of random forest classification algorithms on the full features space (all combined predictor types), as well as on type-specific feature spaces (overlap counts, overlap predictors, and distances) across each resampling technique. It was found that, for 10 kb resolution data, models with distance type predictors had consistently higher AUPRC values compared to other predictor types for both GM12878 and K562 cell lines (Figure 4-8).  This result was observed across all other data resolutions as well (Supplementary Figures 9-23).

@@

The low baseline thresholds represented by the dashed line in the precision-recall curves (Figure 4 and Supplementary Figure 2) indicate the wide disparity between majority and minority classes. Ideally a manageable dataset will have an imbalance ratio hovering near 1. However, in our study we found that there was an average imbalance ratio of 0.154 (SD=0.022) and 0.137 (SD=0.0181) for cell lines GM12878 and K562 respectively (Supplementary Figure 5).  This highlights the need for suitable resampling techniques.

>>>>

We evaluated and compared multiple random forest classification algorithms built on domain data extracted from Hi-C datasets for 2 cell lines C = {GM12878, K562}, at 4 different resolutions R = {10 kb, 25 kb, 50 kb, 100 kb}. For each classifier, we compared 3 predictor types P = {OC, OP, Distance} and 5 class balancing techniques B = {None, RUS, ROS, SMOTE, ROSE}. A system produced from the ensemble framework was defined as: E = {c, r, p, b}, where $c \in C, \quad R \in R, \quad p \in P,$ and $b \in$ B. Letting |X| denote the cardinality of each set, we evaluated $|C| \times |R| \times |P| \times |B|$ different ensemble systems, totaling X (=$x \times 4 \times 3 \times 5$) separate models. For each model, we aimed to assess both performance and identify the genomic annotations that were most influential in correctly labeling the genomic bins that overlapped with flanked TAD boundaries. For simplicity, we present the most significant results, with the complete results available as Supplementary Tables and Figures.

We first examined the class imbalance of the different data resolutions (Supplementary Table ). While the total number of genomic bins decreased as resolution increased, the imbalance ratios remained relatively stable. We found that there was an average imbalance ratio of 0.154 (SD=0.022) and 0.137 (SD=0.0181) for cell lines GM12878 and K562 respectively. The low imbalance ratios indicate a wide disparity between majority and minority classes, and therefore suitable resampling techniques are needed.

@@

For each cell line and resolution, we balanced the data using each resampling technique and performed a series of random forest classification algorithms on the full features space (all combined predictor types), as well as on type-specific feature spaces (overlap counts, overlap predictors, and distances). PR-curves were plotted and compared for 10 kb resolutions (Figures 4-8). For all curves, the baseline threshold used to assign a performance to a random model is defined as the ratio of genomic bins that overlapped with a flanked TAD boundary divided by the total number of genomic bins for the test set, respective to cell line and resolution. It was found that for the 10 kb resolution data, models with distance type predictors had consistently higher AUPRC values compared to other predictor types, across both cell line and resampling technique. This result was observed for the other resolutions as well (Supplementary Figuers 9-23l; Supplementary Table 2).

@@
